---
title: "Exploring GEO Data with GSEexplore"
subtitle: "A walkthrough using GSE63310"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gse63310_walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,  # Set to TRUE if GEO data download works in CI
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

GSEexplore is an interactive Shiny application designed to simplify the exploration of Gene Expression Omnibus (GEO) datasets. This vignette walks through the complete workflow using GSE63310, a publicly available microarray dataset.

The package provides:

- **Data Fetching**: Retrieve GEO datasets directly by accession ID
- **Quality Control**: Assess data quality through PCA and sample distribution analysis
- **Differential Expression**: Perform statistical testing to identify differentially expressed genes
- **Interactive Exploration**: Examine individual gene expression patterns across samples

## Getting Started

### Installation

Install GSEexplore from its repository:

```{r setup, eval=TRUE}
# This would be installed via devtools::install_github() in practice
library(GSEexplore)
```

### About GSE63310

GSE63310 is a microarray study examining gene expression in a biological system with different treatment groups. It contains:

- ~48,000 probes (features)
- Multiple sample replicates
- Defined phenotype/treatment information

## Workflow: Fetching and Exploring Data

### Step 1: Fetch the Dataset

Use `fetch_geo_dataset()` to retrieve the ExpressionSet object from GEO:

```{r fetch}
# Fetch GSE63310
eset <- fetch_geo_dataset("GSE63310")

# Check the object
dim(eset)  # Dimensions: features × samples
```

The function returns a Biobase `ExpressionSet` object containing:
- **Expression matrix**: Gene expression values
- **Phenotype data**: Sample metadata and treatment information
- **Feature data**: Gene/probe annotations

### Step 2: Explore Dataset Summary

Get a quick overview of the dataset:

```{r summary}
summary_info <- get_dataset_summary(eset)

cat("Number of genes/probes:", summary_info$n_features, "\n")
cat("Number of samples:", summary_info$n_samples, "\n")
cat("Phenotype columns:", paste(summary_info$phenotype_cols, collapse=", "), "\n")
cat("Expression range:", summary_info$expression_range[1], "to", summary_info$expression_range[2], "\n")
```

### Step 3: Examine Sample Metadata

```{r metadata}
pdata <- Biobase::pData(eset)
head(pdata)  # View first few rows

# Check available phenotype columns
colnames(pdata)
```

## Quality Control Analysis

### Prepare Data for Analysis

```{r prepare_data}
data_list <- prepare_analysis_data(eset)

# data_list contains:
# - expr_matrix: Expression matrix (genes × samples)
# - phenotypes: Phenotype data
# - feature_names: Gene/probe identifiers
# - sample_names: Sample identifiers
```

### PCA Analysis

Perform Principal Component Analysis to visualize sample relationships:

```{r pca}
pca_result <- compute_pca(data_list$expr_matrix)

# Access results
head(pca_result$scores)  # PC scores for each sample
pca_result$var_explained[1:5]  # % variance explained by first 5 PCs

# The PCA object itself
str(pca_result$pca)
```

### Sample Distribution Analysis

Calculate distribution statistics for each sample:

```{r distributions}
dist_stats <- calculate_sample_distributions(data_list$expr_matrix)
head(dist_stats)

# Columns include: mean, median, sd, quartiles, range, n_genes
```

This helps identify outliers or technical artifacts:

```{r dist_check}
# Check for samples with unusual mean or sd
summary(dist_stats$mean)
summary(dist_stats$sd)
```

## Batch Effect Detection and Correction

### Detect Batch Effects

Check for unwanted variation in your data:

```{r batch_detect}
batch_info <- detect_batch_effects(
  data_list$expr_matrix,
  data_list$phenotypes,
  phenotype_col = "treatment"  # The biological grouping
)

cat("Surrogate variables detected:", batch_info$n_sv, "\n")
cat("Has batch effects:", batch_info$has_batch, "\n")
```

If batch effects are detected, you can correct them:

```{r batch_correct, eval=FALSE}
# Assuming a 'batch' column in phenotype data
batch_variable <- data_list$phenotypes$batch

corrected_matrix <- correct_batch_effects(
  data_list$expr_matrix,
  batch_variable
)

# Use corrected matrix for downstream analysis
```

## Differential Expression Analysis

### Identify Treatment Groups

First, determine the phenotype column for analysis:

```{r de_groups}
# Example: if phenotype data has a "treatment" column
pdata <- Biobase::pData(eset)

# View unique values in a potential group column
# (This depends on your specific dataset structure)
# unique(pdata$treatment)
```

### Run Differential Expression Analysis

Use `run_differential_expression()` to compare two groups:

```{r de_analysis}
# Example: comparing two treatment groups
# Replace "treatment", "control", "treated" with your actual column and values
de_results <- run_differential_expression(
  eset,
  phenotype_col = "treatment",  # The phenotype column
  group1 = "control",           # Reference group
  group2 = "treated"            # Treatment group
)

head(de_results)

# Results include:
# - gene_id: Gene/probe identifier
# - logFC: Log2 fold change
# - AveExpr: Average expression
# - t: t-statistic
# - P.Value: Raw p-value
# - adj.P.Val: Adjusted p-value (Benjamini-Hochberg)
# - significance: * marks for significance levels
```

### Explore Results

```{r de_explore}
# How many genes significant at FDR < 0.05?
sig_genes <- de_results$adj.P.Val < 0.05
cat("Significant genes (FDR < 0.05):", sum(sig_genes), "\n")

# Top genes by adjusted p-value
head(de_results[order(de_results$adj.P.Val), ], 10)

# Genes with largest fold changes
top_fc <- de_results[order(abs(de_results$logFC), decreasing = TRUE), ]
head(top_fc, 10)
```

## Using the Shiny Application

For interactive exploration, launch the Shiny app:

```{r run_app, eval=TRUE}
# Run the interactive Shiny application
# GSEexplore::run_app()
```

The Shiny app provides an intuitive interface with tabs for:

1. **Data Loading**: Enter a GSE accession and fetch data
2. **QC Views**: Interactive PCA and distribution plots
3. **Differential Expression**: Select groups, run analysis, visualize volcano plots
4. **Batch Correction**: Detect and correct for batch effects
5. **Heatmap**: Visualize expression patterns with clustering
6. **Exploration**: Browse individual gene expression across samples

All plots are interactive using plotly, and results tables are searchable and sortable.

## Advanced Usage

## Advanced Usage

### Heatmap Visualization

Create expression heatmaps for exploring gene expression patterns:

```{r heatmap_viz, eval=FALSE}
# Get top variable genes
top_var <- get_top_variable_genes(data_list$expr_matrix, top_n = 100)

# Plot heatmap for top 50 genes
plot_expression_heatmap(
  data_list$expr_matrix,
  data_list$phenotypes,
  gene_indices = top_var$row_index[1:50],
  color_by = "treatment",
  main = "Top 50 Most Variable Genes"
)
```

### Working with Batch-Corrected Data

```{r batch_workflow, eval=FALSE}
# 1. Detect batch effects
batch_info <- detect_batch_effects(
  data_list$expr_matrix,
  data_list$phenotypes,
  "batch_column"
)

if (batch_info$has_batch) {
  # 2. Correct for batch
  corrected <- correct_batch_effects(
    data_list$expr_matrix,
    data_list$phenotypes$batch_column
  )

  # 3. Run analysis on corrected data
  # (would require creating a new ExpressionSet with corrected matrix)
}
```

### Accessing Raw Components

You can work directly with the underlying R functions:

```{r advanced}
# Extract expression matrix
expr_matrix <- Biobase::exprs(eset)
head(expr_matrix[1:5, 1:5])

# Manual PCA
pca_manual <- stats::prcomp(t(expr_matrix), scale = TRUE)
```

### Custom Analysis Pipeline

Combine multiple functions for custom workflows:

```{r custom_pipeline, eval=FALSE}
# 1. Fetch and validate
eset <- fetch_geo_dataset("GSE63310")
GSEexplore:::validate_expressionset(eset)  # Internal validation

# 2. QC assessment
data <- prepare_analysis_data(eset)
pca <- compute_pca(data$expr_matrix)
dist_summary <- calculate_sample_distributions(data$expr_matrix)

# 3. Batch correction if needed
batch_info <- detect_batch_effects(data$expr_matrix, data$phenotypes, "batch")
if (batch_info$has_batch) {
  data$expr_matrix <- correct_batch_effects(data$expr_matrix, data$phenotypes$batch)
}

# 4. Multiple DE comparisons (loop over groups)
# 5. Heatmap visualization
# 6. Integrate with other Bioconductor tools
```

## Common Issues and Solutions

**Issue**: Dataset download fails with network error
- **Solution**: Check internet connection; GEOquery may retry automatically

**Issue**: Too few samples or features for analysis
- **Solution**: Ensure dataset is suitable (typically needs 2+ samples per group, 1000+ features)

**Issue**: Phenotype column has unexpected values
- **Solution**: Examine `pData(eset)` to confirm column names and values before analysis

## Further Reading

- [Biobase Package](https://bioconductor.org/packages/release/bioc/html/Biobase.html)
- [GEOquery Documentation](https://bioconductor.org/packages/release/bioc/html/GEOquery.html)
- [limma User's Guide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)
- [Gene Expression Omnibus](https://www.ncbi.nlm.nih.gov/geo/)

## Session Information

```{r sessionInfo}
sessionInfo()
```
